### 2019-05-01 Paper.Rmd
Final draft before submission

---
title             : "Quantitative involvement moderates the interplay between mothers' versus fathers' qualitative parenting behaviors and children's emotional adjustment development"
shorttitle        : "QUALITATIVE AND QUANTITATIVE INVOLVEMENT"

@@

figsintext        : no
figurelist        : no
tablelist         : no
footnotelist      : no
lineno            : yes
mask              : yes

class             : "man"
output            : papaja::apa6_pdf
---

@@

Emotional adjustment, defined as the absence of internalizing symptoms and manifest signs of emotional dysregulation [@goodmanPsychometricPropertiesStrengths2001], is an important developmental outcome, integral to social competence, academic achievement, and mental health [@grossHandbookEmotionRegulation2013; @aldaoEmotionregulationStrategiesPsychopathology2010]. Fathers have been hypothesized to play a unique role in children's emotional adjustment development, because they encourage children to challenge themselves, in play and in life [@flouriFathersBehaviorsChildren2010; @bogelsDoesFatherKnow2011; @parkeFatherInvolvement2000]. The classical notion of the “attachment relationship” is rooted in security and exploration (REF). Fatherhood researchers have posited that, whereas the mother-child attachment relationship primarily provides warmth and security, exploration might be somewhat more central to the father-child relationship, which can has thus been construed as an “activation relationship” [@paquetteFatherChildActivationRelationship2013]. By encouraging children to venture outside their comfort zone, fathers are thought to stimulate a broad spectrum of emotions, thereby structuring opportunities to expand children's emotional repertoire.

@@

The past decades have seen major strides in the consolidation of theories to guide fatherhood research. Initially, such research was inspired by two societal shifts: Increasing divorce rates, and mothers' greater labor force participation. Both shifts impacted the extent of fathers' involvement in childrearing. Early research focused primarily on the *quantity* of paternal involvement: Whether fathers were present or absent in divorced families, which constitutes a dichotomous operationalization of quantitative involvement, and how much they contributed to the work of childcare in intact families, i.e., a continuous operationalization [@lambBiosocialPerspectivePaternal1987]. Over time, the notion of the father as a coparent gained ground [@pleckFatherhoodIdealsUnited1997], according to which fathers share the responsibilities of childrearing with mothers, and might, to some extent, adopt roles that were historically fulfilled by mothers. To keep pace with changing conceptualizations of fatherhood, research on father involvement evolved to encompass *qualitative* dimensions as well, including warmth and control [or: demandingness; @maccobySocializationContextFamily1983; @pleckPaternalInvolvementRevised2004].

Although it is now widely acknowledged that both the quality and quantity of parental involvement are relevant factors in child development [@faganShouldResearchersConceptualize2014; @pleckPaternalInvolvementRevised2004], these dimensions are rarely jointly investigated. This might be, in part, because research interests have historically varied across discipline lines: Developmental psychology has focused primarily on how qualitative parenting behaviors influence child development, whereas sociology and gender studies have focused more on involvement quantity, and its implications for maternal wellbeing [@craigDoesFatherCare2006; @faganShouldResearchersConceptualize2014]. The present study attempts to bridge these approaches by investigating both qualitative and quantitative involvement in relation to children's emotional adjustment.

### Parenting effects

The standard family environment model posits that the quality of parents' involvement is a proximal mechanism, shaping children's development [@amatoParentalDivorceMarital2008]. Qualitative parenting is typically examined in terms of the constituent dimensions of the parenting styles taxonomy [@baumrindChildCarePractices1967]: Responsiveness and demandingness [@maccobySocializationContextFamily1983]. Responsiveness refers to warm and supportive parenting, which substantial evidence indicates is positively associated with children's socio-emotional adjustment [@khalequePerceivedParentalAcceptanceRejection2002; @morrisRoleFamilyContext2007]. Demandingness encompasses several parenting practices related to setting and enforcing expectations for children's behavior. Two aspects of demandingness investigated in the present study are consistency; the extent to which consequences for incursions are clear and enforced, and inductive reasoning; explaining the reasons for rules and punishments. The literature on links between demandingness and emotional adjustment is more limited, and mixed [@morrisRoleFamilyContext2007]; in part because some forms of demandingness help children meet societal expectations, whereas other forms are overly restrictive, and undermine children's autonomy needs [@barberCentralityControlParenting2013]. A fourth dimension, which has consistently been linked with children's emotional maladjustment, but is distinct from the parenting styles typology, is negative parenting [@changHarshParentingRelation2003; @eisenbergParentalReactionsChildren1999]. We operationalized negative parenting as one of its specific subdimensions: parental anger. The anger parents express in interactions with children might reflect faillure to regulate their own negative emotions, and models the use of anger in communication, and undermines children's emotion regulation [@scaramellaParentingChildrenDistress2008].

<!-- 
Inductive reasoning is thought to be integral to children's development of moral emotions [@HoffmanEmpathymoraldevelopment2000], and might thus play a positive role general emotional development as well.
-->

Although some research suggests that both parents' qualitative parenting behaviors influence children in similar ways [@lambDevelopmentSignificanceFatherChild2004; @faganShouldResearchersConceptualize2014], there is also evidence for unique roles of mothers versus fathers. Fathers' qualitative involvement is associated with emotional adjustment, even after accounting for mothers' influence [@flouriFathersBehaviorsChildren2010]. Furthermore, whether for evolutionary or cultural reasons [@bogelsDoesFatherKnow2011], mothers and fathers may tend to adopt somewhat specialized roles in terms of qualitative parenting: The aforementioned notions of "attachment" and "activation" relationships posit that mothers empathize more with their children, and  provide more warmth and support, whereas fathers activate children by encouraging them to take risks and overcome obstacles, and live up to certain standards for behavior [@majdandzicFathersChallengingParenting2014; @paquetteTheorizingFatherchildRelationship2004]. Empirical research supports these unique roles for mothers and fathers in children's emotional adjustment development: Mothers tend to encourage the expression of emotions, and respond warmly and supportively, whereas fathers structure and sanction emotion expressions to a greater extent [@cassanoInfluenceGenderParental2007; @eisenbergParentalReactionsChildren1999]. Moreover, mothers' supportive behavior was found to be a more salient socialization mechanism than control for emotional adjustment in adolescence, whereas this pattern was reversed for fathers [@vanlissaRoleFathersMothers2018a]. This differentiated pattern was also found with regard to younger children's psychological adjustment [@mcdowellDifferencesMothersFathers2003]. Furthermore, there is some evidence that mothers' negative parenting has a stronger negative impact upon children's emotional adjustment than fathers' negative parenting [@changHarshParentingRelation2003]; perhaps because this behavior is incongruous with mothers' prototypical role as an attachment figure. Based on these findings, we might expect emotional adjustment to be more strongly associated with maternal warmth, and with paternal demanding behaviors, such as consistency and inductive reasoning.

### Child effects

According to the transactional view of family relationships, parents and children reciprocally influence one another [@kerrPerceivedParentingStyle2012; @amatoParentalDivorceMarital2008]. This view of the family as a dynamic system whose members reciprocally influence one another has become widely accepted [@kerrPerceivedParentingStyle2012]. Nevertheless, examining bi-directional effects is not yet standard practice in the childhood literature [@flouriFathersBehaviorsChildren2010]. Available empirical evidence suggests that children's adjustment problems elicit less warmth, more demandingness, and more negative parenting from parents [@gadeyneLongitudinalRelationsParenting2004; @nepplIntergenerationalContinuityParenting2009; @LenguaBidirectionalassociationstemperament2005]. In the adolescent literature, bi-directional effects are more commonly examined. This might be because the notion of children influencing parents is more salient in adolescence, when parent-child relationships become increasingly egalitarian. Findings from the adolescent literature are broadly in line with findings in childhood; suggesting that emotional adjustment problems predicts decreased parental support, and greater control [@OtterpohlCrossLaggedRelationsParenting2015; @VanderGiessenPerceivedAutonomySupport2014; @Wijsbroekdirectioneffectsperceived2011]. Thus, parents' responses to maladjustment appear to be a combination of punitive responses, including decreased support and increased negative parenting, and corrective responses, such as increased demandingness. Moreover, parents' responses to adjustment problems are contingent on children's age: Because younger children are perceived to be less competent than older children, mothers respond to their misbehavior more inductively, and less punitively [@Dixaffectiveorganizationparenting1991]. We might thus expect emotional adjustment to positively predict parental warmth, and negatively predict consistency, inductive reasoning, and anger.

### Assessing the direction of effects

Methods commonly used to examine direction of effects between parents and children have been criticized for yielding biased results if there are stable between-family differences in the variables studied [@Hamakercritiquecrosslaggedpanel2015]. Recent methodological innovations provide better estimates of dynamic reciprocal processes within families, by controlling for stable between-family differences: The random-intercept cross-lagged panel model, which is an extension of the standard cross-lagged panel model [@Hamakercritiquecrosslaggedpanel2015], and dynamic structural equation modeling (DSEM); a multilevel generalization of this model [@AsparouhovDynamicStructuralEquation2018].

Several studies have re-examined the direction of effects between parents and children using these innovative methods, but mostly in adolescent samples. One such study found that predictive effects between parenting and delinquency adjustment that were apparent using classical methods, were found to be explained by correlations between stable between-family differences in parenting and adjustment, once these components were included in the model [@KeijsersParentalmonitoringadolescent2016]. Another study examined the interplay between maternal and paternal qualitative parenting dimensions and adolescents' emotional adjustment, and similarly found that most associations manifested as between-family level correlations. Furthermore, child effects were found to substantially outnumber parenting effects [@vanlissaRoleFathersMothers2018a]. The authors suggested that this might be due to the relative equality of adolescent-parent relationships, as compared to parent-child relationships. If this interpretation is correct, we might expect to find more prominent parenting effects in the present study, with a younger sample. These preliminary findings highlight the fact that the direction of effects between parents and children, while controlling for stable between-family differences, remains an important issue to be examined. At this stage, however, it is difficult to formulate specific hypotheses about directionality, and the level at which associations between parenting and child development should occur.

@@

## Quantitative involvement

### Operationalizing quantitative involvement

Quantitative involvement has been researched extensively in terms of the amount of time spent caring for the child [see @pleckPaternalInvolvementRevised2004]. However, some have argued that the the way childcare tasks are divided is more relevant for children's adjustment than the absolute amount of time invested by each parent [@KrollEarlyFatherInvolvement2016; @sabattiniRelationMothersFathers2004; @craigHowMothersFathers2011; @feinbergInternalStructureEcological2003]. Indeed, research has shown that the equity of the childcare distribution affects child development directly [@deutschPaternalParticipationChild2001], and impacts the emotional climate of the interparental relationship [@CraigDoesFatherCare2006]. When childcare is unequally distributed, women often perform a disproportionate share of the work. This strains the marital relationship [@shockleyItNotWhat2018], and the resulting stress erodes parenting quality, particularly for parents of difficult children [@SheltonInterparentalconflictnegative2008].

There are also methodological advantages to focusing on the distribution of childcare. Operationalizing quantitative involvement in terms of both parents' absolute involvement, e.g., as hours spent caring for the child, incurs a potential confound: Any effect that appears to be due to fathers' greater involvement might, in fact, be due to children receiving greater a greater total amount of parental care [see @deutschPaternalParticipationChild2001]. This confound is mitigated by operationalizing quantitative involvement in terms of the childcare distribution, and controlling for total joint care. Some prior research has attempted to capture the childcare distribution directly, by asking which has most responsibility for certain tasks, on a continuum from the mother doing everything to the father doing everything [@KrollEarlyFatherInvolvement2016]. When quantitative involvement is measured in terms of absolute involvement, however, an elegant statistical solution can disentangle the childcare distribution from total joint care. The so-called "latent mean and difference" approach, which is typically used to capture the difference between consecutive longitudinal measurements [@McArdleLatentVariableModeling2009], can be utilized to re-express mothers' and fathers' absolute quantitative involvement into a "latent mean" of involvement, which represents the total joint care provided by both parents, and latent differential involvement, which represents the extent to which one parent is more involved than the other. This approach taps directly into a theoretically relevant construct, and prevents confounding. 
## Main effects of quantitative involvement

@@

By the same logic, we might expect that, when one parent is more involved than the other, the parenting quality of the more involved parent would improve. However, an argument to the contrary can be made as well. Retrospective reports by 19-year old participants indicate that an unequal childcare distribution tends to go hand in hand with suboptimal qualitative parenting behaviors - specifically, increased authoritarian parenting in mothers, and increased disengaged parenting in fathers [@sabattiniRelationMothersFathers2004]. Greater differential involvement might undermine both parents' involvement quality, through its effect on the marital relationship. Childcare responsibilities are a heavy burden if they come down on a single parent's shoulders [@deutschPaternalParticipationChild2001]. Mothers thus benefit directly when fathers relieve some of this burden [@PleckPaternalinvolvementRevised2004]. Consequently, mothers tend to be more satisfied when fathers are more involved [@PharesFathersMothersInvolvement2009]. An unfair distribution of childcare, on the other hand, is associated with greater maternal anger and resentment [@RossGenderParenthoodAnger1996]. This strains the interparental relationship, interferes with sensitive parenting, and undermines child wellbeing [@davies2006interparental]. In other words, when care is distributed relatively equally, the entire family system will likely be more harmonious. We might thus expect greater differential involvement to aversely affect both parents' parenting quality, as well as children's emotional adjustment.

@@

Although the quantity of paternal involvement has received substantial interest in the fields of sociology and gender studies, this literature has focused primarily on its implications for mothers' time use and wellbeing. Consequences for children's emotional outcomes have received less attention to date. Bioecological theories of development indicate that children require a certain amount of care, and who provides this care is, to some extent, flexible [see @BronfenbrennerEcologyHumanDevelopment1979]. This suggests that the total amount of care provided should be positively associated with children's emotional adjustment. Differential involvement, on the other hand, might be negatively linked to emotional adjustment: When children perceive that one of their parents is relatively uninvolved, they likely feel less supported than when both parents are equally invested in their wellbeing [@deutschPaternalParticipationChild2001].

@@

## Interactions between involvement quality and quantity

Interactions between involvement quality and quantity have largely been overlooked in the aforementioned theoretical models of father involvement; perhaps in part because these two dimensions have rarely been jointly investigated [@JaffeeLifeFatherBenefits2003]. However, It is conceivable that children's emotional adjustment might be differently impacted by paternal control if that father is generally highly involved with the day-to-day minutiae of childrearing, than when he is generally absent. Conversely, children's emotional adjustment might elicit different qualitative parenting behaviors from hands-on fathers, who are aware of the reasons for their children's emotional troubles, than from out-of-touch fathers, who might perceive maladjustment as a problem behavior that should be sanctioned. To capture patterns like these, a model should account for interactions between involvement quality and quantity.

In this respect, theories of father involvement can be complemented by Feinberg's coparenting model [@FeinbergInternalStructureEcological2003]. Coparenting refers to the ways parents relate to each other in the role of parent. Although the coparenting literature is typically invoked to describe childrearing in divorced families, it also holds insights for intact families. One dimension of coparenting is involvement quantity, which Feinberg defines as the "division of labor" in childcare. Note that this definition parallels our operationalization of quantitative involvement. The division of childcare labor is proposed to affect children's adjustment and parents' qualitative childrearing behaviors directly, and to interact with the link between these two constructs.

There is some preliminary empirical support for interactions between involvement quality and quantity; for instance, fathers' playful and caring involvement predicted preschoolers' emotional adjustment in school, but only when accompanied by a supportive coparenting relationship [@JiaTransactionalRelationsFather2012]. Furthermore, in adolescence, the quantity of paternal education-related structure, an age-appropriate form of involvement, interacted with paternal warmth to predict both academic achievement and behavioral adjustment [@WangParentalInvolvementAfrican2014]. Applying this insight from the coparenting literature, we might thus expect parents' quantitative involvement to be directly associated with their qualitative parenting behaviors and children's emotional adjustment, and to moderate effects between qualitative parenting behaviors and emotional adjustment.

@@

The present study aimed to make several contributions to the literature on fathers' role in their children's emotional adjustment development. First, we included both mothers and fathers, in order to tease apart their common and unique roles. Second, our operationalization of involvement incorporated both qualitative involvement, in terms of specific parenting behaviors, and quantitative involvement, operationalized as total joint childcare and differential involvement. Third, we employed recent methodological advances to disentangle stable between-family differences from dynamic within-family predictive processes, in order to more accurately assess the direction of effects between qualitative parenting behaviors and emotional adjustment. Fourth, we examined interactions between qualitative and quantitative involvement, which have been largely overlooked in prior work.

Based on our review of the literature, we expected specific dimensions of qualitative involvement to be associated with emotional adjustment at both the within- and between-family level. The literature suggests that warmth, consistency, and inductive reasoning should be positively associated, whereas negative parenting should be negatively associated, with emotional adjustment. With regard to child effects, we might expect some of these paths to be reversed: Positive adjustment might lead to diminished inductive reasoning and parental anger. We further hypothesized that quantitative involvement would predict children's adjustment directly: Greater joint total care should predict better emotional adjustment, whereas a more skewed distribution of care should predict poorer emotional adjustment. Greater joint care should also predict both parents' qualitative involvement, in the sense of lower parental anger, and greater consistency, inductive reasoning, and warmth. A more skewed distribution of care, on the other hand, should show these associations only for the more involved parent. Importantly, we also hypothesized that involvement quality and quantity would interact, such that a more skewed distribution of quantitative involvement would interfere with the effects of qualitative parenting behaviors on children, and parents' adaptive responses to children's maladjustment. In other words, we expected that an unfair childcare distribution would predict suboptimal parenting, and parental responses to maladjustment, due to the strain such a distribution places upon the entire family system.  Finally, we accounted for potential effects of child sex.

@@

Measures of quantitative involvement varied from wave to wave, due to children's changing care needs. We used reports at T1, when both parents rated their involvement in four different child care activities on a five-point scale ranging from "once a day or more" to "not at all" (e.g.: "In the past month how often did you help this child brush his/her teeth?").

@@

<!-- structure notes
1) mothers and fathers, in order to tease apart their common and unique roles. 
2) incorporated quantitative involvement, in terms of total joint childcare and the childcare distribution, and qualitative involvement
3) interactions between qualitative and quantitative involvement
4) assess the direction of effects between parenting behaviors and emotional adjustment.

>>>>

---
title             : "Quantitative involvement moderates mothers' versus fathers' roles in children's emotional adjustment development"
shorttitle        : "QUANTITATIVE AND QUALITATIVE INVOLVEMENT"

@@

figsintext        : no
figurelist        : no
tablelist         : no
footnotelist      : no
lineno            : yes
mask              : no

class             : "man"
output            : papaja::apa6_word
---

@@

Fathers have been hypothesized to play a unique role in children's emotional adjustment development, because they encourage children to challenge themselves, in play and in life [@FlouriFathersbehaviorschildren2010; @BogelsDoesFatherKnow2011; @ParkeFatherinvolvement2000]. Emotional adjustment, defined as the absence of internalizing symptoms and manifest signs of emotional dysregulation [@GoodmanPsychometricpropertiesstrengths2001], is an important developmental outcome, because it is integral to social competence, academic achievement, and mental health [@GrossHandbookemotionregulation2013; @AldaoEmotionregulationstrategiespsychopathology2010]. By encouraging children to venture outside their comfort zone, fathers are thought to stimulate a broad spectrum of emotions, thereby structuring opportunities to expand children's emotional repertoire. The classical notion of the “attachment relationship” is rooted in security and exploration (REF). Fatherhood researchers have posited that, whereas the mother-child attachment relationship primarily provides warmth and security, exploration might be somewhat more central to the father-child relationship, which can has thus been construed as an “activation relationship” [@PaquetteFatherChildActivationRelationship2013].

@@

The past decades have seen major strides in the consolidation of theories to guide fatherhood research. Initially, such research was inspired by two societal shifts: Increasing divorce rates, and mothers' greater labor force participation. Both shifts impacted the extent of fathers' involvement in childrearing. Early research focused primarily on the *quantity* of paternal involvement: Whether fathers were present or absent in divorced families, which constitutes a dichotomous operationalization of quantitative involvement, and how much they contributed to the work of childcare in intact families, i.e., a continuous operationalization [@Lambbiosocialperspectivepaternal1987]. Over time, the notion of the father as a coparent gained ground [@PleckFatherhoodidealsUnited1997], according to which fathers share the responsibilities of childrearing with mothers, and might, to some extent, adopt roles that were historically fulfilled by mothers. To keep pace with changing conceptualizations of fatherhood, research on father involvement evolved to encompass *qualitative* dimensions as well, including warmth and control [or: demandingness; @MaccobySocializationcontextfamily1983; @PleckPaternalinvolvementRevised2004]. Nowadays, it is widely acknowledged that both the quantity and quality of parental involvement are relevant factors in child development [@FaganShouldresearchersconceptualize2014; @PleckPaternalinvolvementRevised2004]. Interactions between quantity and quality have rarely been considered, however, perhaps in part because, historically, interest in these dimensions has varied across discipline lines. Sociology and gender studies, for instance, have primarily focused on involvement quantity [@CraigDoesFatherCare2006; @FaganShouldresearchersconceptualize2014], whereas developmental psychology has focused on how qualitative parenting behaviors influence child development. The present study attempts to bridge these disciplines, by investigating whether the division of quantitative involvement between mothers and fathers moderates the interplay between qualitative parenting behaviors and children's emotional adjustment.

## Interactions between involvement quantity and quality

The aforementioned theoretical models of fatherhood are limited, in the sense that they have underemphasized potential interactions between involvement quantity and quality [@JaffeeLifeFatherBenefits2003]. For example, children's emotional adjustment might be differently impacted by a father's efforts to exert control if he is generally highly involved with the day-to-day minutiae of childrearing, than when he is generally absent. Conversely, children's emotional adjustment might elicit different qualitative parenting behaviors from hands-on fathers, who are aware of the reasons for their children's emotional troubles, than from out-of-touch fathers, who might perceive maladjustment as a problem behavior that should be sanctioned. To capture these patterns, a model should account for interactions between involvement quantity and quality.

With regard to such interactions, existing theories of father involvement are complemented by Feinberg's coparenting model [@FeinbergInternalStructureEcological2003]. Coparenting refers to the ways parents relate to each other in the role of parent. Although the coparenting literature is typically invoked to describe childrearing in divorced families, it also holds insights for intact families. One dimension of coparenting is involvement quantity, which Feinberg defines as the "division of labor" in childcare. This dimension is proposed to affect children's adjustment and parents' qualitative childrearing behaviors directly, and to moderate the link between these two constructs.

There is some preliminary empirical support for such interactions between involvement quantity and quality. For instance, fathers' playful and caring involvement predicted preschoolers' emotional adjustment in school, but only when accompanied by a supportive coparenting relationship [@JiaTransactionalRelationsFather2012]. Furthermore, in adolescence, the quantity of paternal education-related structure, an age-appropriate form of involvement, interacted with paternal warmth to predict both academic achievement and behavioral adjustment [@WangParentalInvolvementAfrican2014]. Applying this insight from the coparenting literature, we might thus expect parents' quantitative involvement to be directly associated with their qualitative parenting behaviors and children's emotional adjustment, and to moderate effects between qualitative parenting behaviors and emotional adjustment.

## Operationalizing quantitative involvement

A second important insight from coparenting theory is this notion of the "division of labor", pertaining to childcare. This notion suggests that parents' relative contributions to childrearing might be more relevant for predicting children's adjustment than parent's absolute quantitative involvement [@KrollEarlyFatherInvolvement2016; @sabattiniRelationMothersFathers2004]. Indeed, research has shown that the equity of the childcare labor distribution affects child development directly [@DeutschPaternalParticipationChild2001], and impacts the emotional climate of the interparental relationship [@CraigDoesFatherCare2006]. When childcare is unequally distributed, women often perform a disproportionate share of the work. This strains the marital relationship [@ShockleyItnotwhat2018], and the resulting stress erodes parenting quality, particularly for difficult children [@SheltonInterparentalconflictnegative2008].

There are also methodological advantages to focusing on the distribution of childcare. It has been noted that operationalizing quantitative involvement in terms of both parents' absolute involvement, e.g., as hours spent caring for the child, incurs a potential confound: Any effect that appears to be due to fathers' greater involvement might, in fact, be due to children receiving greater a greater total amount of parental care [see @DeutschPaternalParticipationChild2001]. This confound is mitigated by operationalizing quantitative involvement in terms of the childcare distribution, while controlling for total joint care. Some prior research has attempted to capture the childcare distribution; for example, by asking who has most responsibility for certain task, on a continuum from the mother doing everything to the father doing everything [@KrollEarlyFatherInvolvement2016]. When quantitative involvement has been measured in terms of absolute involvement, there is an elegant statistical solution to disentangle the childcare distribution from total joint care. The so-called "latent mean and difference" approach, which is typically used to capture the difference between consecutive longitudinal measurements [@McArdleLatentVariableModeling2009], can be utilized to re-express mothers' and fathers' absolute quantitative involvement into a "latent mean" of involvement, which represents the total joint care provided by both parents, and latent differential involvement, which represents the extent to which one parent is more involved than the other. This approach taps directly into a theoretically relevant construct, and prevents confounding.

@@

## Main effects of quantitative involvement

@@

By the same logic, we might expect that, when one parent is more involved than the other, the parenting quality of the more involved parent would improve. However, an argument to the contrary can be made as well. Retrospective reports by 19-year old participants indicate that an unequal childcare distribution tends to go hand in hand with suboptimal qualitative parenting behaviors - specifically, increased authoritarian parenting in mothers, and increased disengaged parenting in fathers [@sabattiniRelationMothersFathers2004]. Greater differential involvement might undermine both parents' involvement quality, through its effect on the marital relationship. Childcare responsibilities are a heavy burden if they come down on a single parent's shoulders [@DeutschPaternalParticipationChild2001]. Mothers thus benefit directly when fathers relieve some of this burden [@PleckPaternalinvolvementRevised2004]. Consequently, mothers tend to be more satisfied when fathers are more involved [@PharesFathersMothersInvolvement2009]. An unfair distribution of childcare, on the other hand, is associated with greater maternal anger and resentment [@RossGenderParenthoodAnger1996]. This strains the interparental relationship, interferes with sensitive parenting, and undermines child wellbeing [@davies2006interparental]. In other words, when care is distributed relatively equally, the entire family system will likely be more harmonious. We might thus expect greater differential involvement to aversely affect both parents' parenting quality, as well as children's emotional adjustment.

@@

Although the quantity of paternal involvement has received substantial interest in the fields of sociology and gender studies, this literature has focused primarily on its implications for mothers' time use and wellbeing. Consequences for children's emotional outcomes have received less attention to date. Bioecological theories of development indicate that children require a certain amount of care, and who provides this care is, to some extent, flexible [see @BronfenbrennerEcologyHumanDevelopment1979]. This suggests that the total amount of care provided should be positively associated with children's emotional adjustment. Differential involvement, on the other hand, might be negatively linked to emotional adjustment: When children perceive that one of their parents is relatively uninvolved, they will likely feel less supported than when both parents are equally invested in their wellbeing [@DeutschPaternalParticipationChild2001].

@@

The standard family environment model posits that the quality of parents' involvement is a proximal mechanism, shaping children's development [@AmatoParentalDivorceMarital2008]. Qualitative parenting has often been examined in terms of the dimensions underlying parenting styles [@BaumrindChildcarepractices1967]: Responsiveness and demandingness [@MaccobySocializationcontextfamily1983]. Responsiveness refers to warm and supportive parenting, which substantial evidence indicates is positively associated with children's socio-emotional adjustment [@KhalequePerceivedParentalAcceptanceRejection2002; @Morrisrolefamilycontext2007]. Demandingness encompasses several parenting practices related to setting and enforcing expectations for children's behavior. Two aspects of demandingness investigated in the present study are consistency; the extent to which consequences for incursions are clear and enforced, and inductive reasoning; explaining the reasons for rules and punishments. The literature on links between demandingness and emotional adjustment is more limited, and mixed [@Morrisrolefamilycontext2007]; in part because some forms of demandingness help children meet societal expectations, whereas other forms are overly restrictive, and undermine children's autonomy needs [@Barbercentralitycontrolparenting2013]. A fourth dimension which has consistently been linked with children's emotional maladjustment - but is distinct from the parenting styles typology - is negative parenting, operationalized here as one of its constitutuent components: parental anger [@ChangHarshParentingRelation2003; @EisenbergParentalReactionsChildren1999]. The anger parents express in interactions with children might reflect faillure to regulate their own negative emotions, and models the use of anger in communication, and undermines children's emotion regulation [@ScaramellaParentingChildrenDistress2008].

<!-- 
Inductive reasoning is thought to be integral to children's development of moral emotions [@HoffmanEmpathymoraldevelopment2000], and might thus play a positive role general emotional development as well.

-->

Although some research suggests that both parents' qualitative parenting behaviors influence children in similar ways [@LambDevelopmentSignificanceFatherChild2004; @FaganShouldresearchersconceptualize2014], there is also evidence for unique roles of mothers versus fathers. Fathers' qualitative involvement is associated with emotional adjustment, even after accounting for mothers' influence [@FlouriFathersbehaviorschildren2010]. Furthermore, whether for evolutionary or cultural reasons [@BogelsDoesFatherKnow2011], mothers and fathers may tend to adopt somewhat specialized roles in terms of qualitative parenting: The aforementioned notions of "attachment" and "activation" relationships posit that mothers empathize more with their children, and  provide more warmth and support, whereas fathers activate children by encouraging them to take risks and overcome obstacles, and live up to certain standards for behavior [@MajdandzicFathersChallengingParenting2014; @PaquetteTheorizingfatherchildrelationship2004]. Empirical research supports these unique roles for mothers and fathers in children's emotional adjustment development: Mothers tend to encourage the expression of emotions, and respond warmly and supportively, whereas fathers structure and sanction emotion expressions to a greater extent [@CassanoInfluenceGenderParental2007; @EisenbergParentalReactionsChildren1999]. Moreover, mothers' supportive behavior was found to be a more salient socialization mechanism than control for emotional adjustment in adolescence, whereas this pattern was reversed for fathers [@VanLissarolefathersmothersinpress]. This differentiated pattern was also found with regard to younger children's psychological adjustment [@McDowellDifferencesMothersFathers2003]. Furthermore, there is some evidence that mothers' negative parenting has a stronger negative impact upon children's emotional adjustment than fathers' negative parenting [@ChangHarshParentingRelation2003]; perhaps because this behavior is incongruous with mothers' prototypical role as an attachment figure. Based on these findings, we might expect emotional adjustment to be more strongly associated with maternal warmth, and with paternal demanding behaviors, such as consistency and inductive reasoning.

## Child effects

One major limitation of correlational research is that associations between parenting and children's adjustment might be explained equally well by child effects on parents. There is now substantial evidence to suggest that children's adjustment problems elicit less warmth, more demandingness, and more negative parenting from parents [@GadeyneLongitudinalRelationsParenting2004; @NepplIntergenerationalContinuityParenting2009; @LenguaBidirectionalassociationstemperament2005]. Child effects have been more consistently investigated in the adolescent literature than in the childhood literature. This might be because the notion of children influencing parents is more salient in adolescence, when parent-child relationships become increasingly egalitarian. Nonetheless, findings from the adolescent literature are broadly in line with the childhood literature, suggesting that emotional adjustment problems predicts decreased parental support, and greater control [@OtterpohlCrossLaggedRelationsParenting2015; @VanderGiessenPerceivedAutonomySupport2014; @Wijsbroekdirectioneffectsperceived2011]. Thus, parents' responses to maladjustment appear to be a combination of punitive responses, including decreased support and increased negative parenting, and corrective responses, such as increased demandingness. Moreover, parents' responses to adjustment problems are contingent on children's age: Because younger children are perceived to be less competent than older children, mothers respond to their misbehavior more inductively, and less punitively [@Dixaffectiveorganizationparenting1991]. We might thus expect emotional adjustment to positively predict parental warmth, and negatively predict consistency, inductive reasoning, and anger.

According to the transactional view of family relationships, parents and children reciprocally influence one another [@KerrPerceivedparentingstyle2012; @AmatoParentalDivorceMarital2008]. This view of the family as a dynamic system whose members reciprocally influence one another is now widely accepted [@KerrPerceivedparentingstyle2012]. Nevertheless, it is not yet common practice to examine bi-directional effects [@FlouriFathersbehaviorschildren2010]. Moreover, methods commonly used to examine direction of effects have recently been criticized for yielding biased results if there are stable between-family differences in the variables studied [@Hamakercritiquecrosslaggedpanel2015]. Recent methodological innovations provide better estimates of dynamic reciprocal processes within families, by controlling for stable between-family differences: The random-intercept cross-lagged panel model, which is an extension of the standard cross-lagged panel model [@Hamakercritiquecrosslaggedpanel2015], and dynamic structural equation modeling (DSEM); a multilevel generalization of this model [@AsparouhovDynamicStructuralEquation2018].

Several studies have re-examined the direction of effects between parents and children using these innovative methods, but mostly in adolescent samples. One such study found that predictive effects between parenting and delinquency adjustment that were apparent using classical methods, were found to be explained by correlations between stable between-family differences in parenting and adjustment, once these components were included in the model [@KeijsersParentalmonitoringadolescent2016]. Another study examined the interplay between maternal and paternal qualitative parenting dimensions and adolescents' emotional adjustment, and similarly found that most associations manifested as between-family level correlations. Furthermore, child effects were found to substantially outnumber parenting effects [@VanLissarolefathersmothersinpress]. The authors suggested that this might be due to the relative equality of adolescent-parent relationships, as compared to parent-child relationships. If this interpretation is correct, we might expect to find more prominent parenting effects in the present study, with a younger sample. These preliminary findings highlight the fact that the direction of effects between parents and children, while controlling for stable between-family differences, remains an important issue to be examined. At this stage, however, it is difficult to formulate specific hypotheses about directionality, and the level at which associations between parenting and child development should occur.

## Differences between sons and daughters

@@

1) Interactions
2) Effect of QUANT
3) Effect of QUAL
4) Direction of effects
5) Sex differences

(1)	Qual effects
(2)	Quant effects
(3)	Interactions
(4)	Child effects
(5)	Direction of effects
(6)	Sex differences

The present study aimed to make several contributions to the literature on fathers' role in their children's emotional adjustment development. First, we included both mothers and fathers, in order to tease apart their common and unique roles. Second, our operationalization of involvement incorporated both quantitative dimensions, in terms of total joint childcare and differential involvement, and qualitative dimensions, in terms of specific parenting behaviors. Third, we examined interactions between quantitative and qualitative involvement, which have been largely overlooked in prior work. Fourth, we employed recent methodological advances to disentangle stable between-family differences from dynamic within-family predictive processes, in order to more accurately assess the direction of effects between qualitative parenting behaviors and emotional adjustment.

Based on our review of the literature, we expected that involvement quantity and quality would interact, such that a more skewed distribution of quantitative involvement would interfere with the effects of qualitative parenting behaviors on children, and parents' adaptive responses to children's maladjustment. In other words, we expected that an unfair childcare distribution would predict suboptimal parenting, and parental responses to maladjustment, due to the strain such a distribution places upon the entire family system. We further hypothesized that quantitative involvement would predict children's adjustment directly: Greater joint total care should predict better emotional adjustment, whereas a more skewed distribution of care should predict poorer emotional adjustment. Greater joint care should also predict both parents' qualitative involvement, in the sense of lower parental anger, and greater consistency, inductive reasoning, and warmth. A more skewed distribution of care, on the other hand, should show these associations only for the more involved parent. Finally, we expected specific qualitative parenting dimensions to be associated with emotional adjustment at both the within- and between-family level. The literature suggests that warmth, consistency, and inductive reasoning should be positively associated, whereas negative parenting should be negatively associated, with emotional adjustment. With regard to child effects, we might expect some of these paths to be reversed: Positive adjustment might lead to diminished inductive reasoning and parental anger. We also accounted for potential effects of child sex.

@@

Due to children's changing care needs and increasing independence, the operational definition of quantitative involvement varied from wave to wave. We therefore used reports at T1 as an indicator of between-family differences in parents' quantitative involvement. Both parents rated their involvement in four different child care activities on a five-point scale ranging from "once a day or more" to "not at all" (e.g.: "In the past month how often did you help this child brush his/her teeth?").

@@

<!-- structure notes
1) mothers and fathers, in order to tease apart their common and unique roles. 
2) incorporated quantitative involvement, in terms of total joint childcare and the childcare distribution, and qualitative involvement
3) interactions between quantitative and qualitative involvement
4) assess the direction of effects between parenting behaviors and emotional adjustment.
